---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Multi-component Bayesian Endpoints (MBEs)

<!-- badges: start -->
[![R-CMD-check](https://github.com/hyejung0/MBE/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hyejung0/MBE/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This repository contains the R package accompanying the
*Multi-Component Bayesian Endpoints for Randomized Clinical Trials* manuscript
by Lee et al. (2026).

This package implements an MBE with exactly one clinical endpoint and two
surrogate endpoints. The current CKD implementation uses chronic and acute
eGFR slopes as the two surrogates. A Bayesian framework incorporates historical
randomized clinical trials (RCTs) and propagates uncertainty when estimating the
treatment effect on the clinical endpoint from both surrogate effects.

See [Mapping the MBE Manuscript to the
Package](https://hyejung0.github.io/MBE/articles/MBE-manuscript-to-code.html)
for the complete workflow.

## Installation

You can install the development version of MBE from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("hyejung0/MBE")
```

Fitting a historical model requires a working CmdStan installation. If CmdStan
has not already been installed, run:

``` r
cmdstanr::check_cmdstan_toolchain()
cmdstanr::install_cmdstan()
```

## Example

The package includes compatible historical posterior draws and simulated data.
The input for a new trial always contains the clinical endpoint, both surrogate
endpoints, and the three pairwise correlations:

```{r example, message=FALSE}
library(MBE)

data("historical_posterior")
data("trial_sim_dat")

#Construct a list for the new trial's observed data, using the first row of the simulated dataset.
new_trial <- list(
  ClnEst = trial_sim_dat$CE_est[1],
  ClnSE = trial_sim_dat$CE_se[1],
  Sur1Est = trial_sim_dat$Sur1_est[1],
  Sur1SE = trial_sim_dat$Sur1_se[1],
  Sur2Est = trial_sim_dat$Sur2_est[1],
  Sur2SE = trial_sim_dat$Sur2_se[1],
  R1Clin = trial_sim_dat$Cor_CE_Sur1[1],
  R2Clin = trial_sim_dat$Cor_CE_Sur2[1],
  R12 = trial_sim_dat$Cor_Sur1_Sur2[1]
)

#Estimate the posterior distribution of the MBE for the new trial using the historical posterior draws.
set.seed(1)
fit <- MBE(historical_posterior, new_trial)

#Mean of the posterior distribution for the clinical endpoint, surrogate 1, and surrogate 2.
fit$post_mean

#Corresponding variance matrix
fit$post_var
```


A detailed example is provided in the package vignette [Simulation Example](https://hyejung0.github.io/MBE/articles/simulation-example.html).


<!-- ## Historical-model assessment -->


<!-- Users fit each historical training set themselves with -->
<!-- `historical_model_fit_2surrogates()`, choosing their intercept and uncertainty -->
<!-- prior. Assessment is separate: `loo_cv_rmse()` accepts a two-column table named -->
<!-- `observed` and `estimated`, and `tolerance_interval_coverage()` calls -->
<!-- `clinical_predictive_distribution()` on existing posterior draws to check -->
<!-- user-selected quantile intervals, such as `probs = c(0.025, 0.975)`. -->

<!-- Each held-out trial serves as a pseudo-new trial in internal cross-validation. -->
<!-- Results for the simulated package dataset are not expected to match the -->
<!-- manuscript's CKD trial results. The manuscript-to-package article demonstrates -->
<!-- the fitting loop, prediction, and aggregation of interval-capture indicators. -->

<!-- The earlier completed 66-trial assessment is retained as two archived datasets. -->
<!-- The RMSE can be recalculated directly from its paired effects. Archived -->
<!-- coverage predates the current prediction and interval-checking method; see the -->
<!-- dataset documentation before comparing it with new coverage calculations. -->

<!-- ```{r loo-results} -->
<!-- data("loo_assessment_summary", package = "MBE") -->
<!-- data("loo_assessment_by_trial", package = "MBE") -->

<!-- loo_assessment_summary -->
<!-- head(loo_assessment_by_trial) -->
<!-- loo_cv_rmse(data.frame( -->
<!--   observed = loo_assessment_by_trial$observed_clinical, -->
<!--   estimated = loo_assessment_by_trial$posterior_mean_clinical -->
<!-- )) -->
<!-- ``` -->
