---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Multi-component Bayesian Endpoints (MBEs)

<!-- badges: start -->
[![R-CMD-check](https://github.com/hyejung0/MBE/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hyejung0/MBE/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This repository contains all relevant R codes that was used in the
Multi-Component Bayesian Endpoints for Randomized Clinical Trials paper
produced by Lee et. al., (2026).

This package implements an MBE with exactly one clinical endpoint and two
surrogate endpoints. The current CKD implementation uses chronic and acute
eGFR slopes as the two surrogates. A Bayesian framework incorporates historical
randomized clinical trials (RCTs) and propagates uncertainty when estimating the
treatment effect on the clinical endpoint from both surrogate effects.

Detailed explanation of mapping the manuscript to the code can be found in this document: [Mapping the manuscript to the code](articles/MBE-manuscript-to-code.html)

## Installation

You can install the development version of MBE from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("hyejung0/MBE")
```

## Example

The package includes compatible historical posterior draws and simulated data.
The input for a new trial always contains the clinical endpoint, both surrogate
endpoints, and the three pairwise correlations:

```{r example, message=FALSE}
library(MBE)

data("historical_posterior")
data("trial_sim_dat")

new_trial <- list(
  ClnEst = trial_sim_dat$CE_est[1],
  ClnSE = trial_sim_dat$CE_se[1],
  Sur1Est = trial_sim_dat$Sur1_est[1],
  Sur1SE = trial_sim_dat$Sur1_se[1],
  Sur2Est = trial_sim_dat$Sur2_est[1],
  Sur2SE = trial_sim_dat$Sur2_se[1],
  R1Clin = trial_sim_dat$Cor_CE_Sur1[1],
  R2Clin = trial_sim_dat$Cor_CE_Sur2[1],
  R12 = trial_sim_dat$Cor_Sur1_Sur2[1]
)

set.seed(1)
fit <- MBE(historical_posterior, new_trial)
fit$post_mean
```
