---
title: "Mapping the MBE Manuscript to the Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping the MBE Manuscript to the Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MBE)
data("trial_sim_dat", package = "MBE")
```

## Overview

The Multi-Component Bayesian Endpoint (MBE) workflow described by Lee et al.
(2026) has two stages:

1. Fit a trial-level Bayesian meta-analytic model to historical randomized
   clinical trials.
2. Use draws from that historical model to update the treatment effects for a
   new trial.

The package implementation is deliberately restricted to one clinical endpoint
and exactly two surrogate endpoints. In the chronic kidney disease example, the
two surrogates are treatment effects on chronic and acute eGFR slopes.

## Input data

`historical_model_fit_2surrogates()` expects one row per historical trial and
the following nine columns:

| Column | Meaning |
|:--|:--|
| `CE_est`, `CE_se` | Clinical treatment-effect estimate and standard error |
| `Sur1_est`, `Sur1_se` | Surrogate 1 estimate and standard error |
| `Sur2_est`, `Sur2_se` | Surrogate 2 estimate and standard error |
| `Cor_CE_Sur1` | Sampling correlation between the clinical endpoint and surrogate 1 |
| `Cor_CE_Sur2` | Sampling correlation between the clinical endpoint and surrogate 2 |
| `Cor_Sur1_Sur2` | Sampling correlation between the two surrogates |

`trial_sim_dat` provides a shareable simulated example with 66 trials. The
restricted CKD-EPI CT trial data used in the motivating analysis are not
distributed with the package.

## Stage 1: fit the historical model

The historical model can use a fixed or estimated clinical intercept and either
half-normal or inverse-gamma priors for its uncertainty parameters. The
manuscript's initial specification uses an estimated (random) intercept and
inverse-gamma priors. The first simulated trial can be held out as follows:

```{r historical-fit, eval=FALSE}
historical_fit <- historical_model_fit_2surrogates(
  data = trial_sim_dat[-1, ],
  random_intercept = TRUE,
  prior_for_uncertainty = "inverse_gamma",
  nchains = 4,
  ncores = 4,
  niter = 2000,
  nwarmup = 1000,
  seed = 2026,
  adapt_delta = 0.95,
  max_treedepth = 15
)

historical_fit$summary
historical_fit$loo
historical_fit$waic
```

The returned object also contains the CmdStan fit, posterior draws, and sampler
diagnostics. Historical fits should be accepted only after checking convergence
and sampler warnings.

## Stage 2: update the MBE for a new trial

The package includes `historical_posterior`, a compact set of compatible draws
from the random-intercept, inverse-gamma model with trial 1 held out. The held-out
trial is converted to the naming convention used by `MBE()`:

```{r mbe-update}
data("historical_posterior", package = "MBE")

new_trial <- list(
  ClnEst = trial_sim_dat$CE_est[1],
  ClnSE = trial_sim_dat$CE_se[1],
  Sur1Est = trial_sim_dat$Sur1_est[1],
  Sur1SE = trial_sim_dat$Sur1_se[1],
  Sur2Est = trial_sim_dat$Sur2_est[1],
  Sur2SE = trial_sim_dat$Sur2_se[1],
  R1Clin = trial_sim_dat$Cor_CE_Sur1[1],
  R2Clin = trial_sim_dat$Cor_CE_Sur2[1],
  R12 = trial_sim_dat$Cor_Sur1_Sur2[1]
)

set.seed(1)
mbe_fit <- MBE(
  mcmc_dat = historical_posterior,
  sample_dat = new_trial,
  diffuse = TRUE,
  diffuse_se = 100,
  intercept0 = TRUE
)

mbe_fit$post_mean
mbe_fit$post_var
mbe_fit$post_quantiles_clinical
mbe_fit$importance_diagnostics
```

`post_mean` and `post_var` are the mean and covariance of the updated posterior
mixture. `post_psi0` contains resampled posterior draws in the fixed order
clinical endpoint, surrogate 1, and surrogate 2. `weight_data$norm_w` contains
the normalized importance weights. `importance_diagnostics` helps identify
whether those weights are concentrated in only a small number of historical
draws.

## Historical-model fit assessment

The historical model is assessed by holding out each of the 66 trials in turn.
`fit_loo_historical_models()` fits the random-intercept, inverse-gamma model to
the other 65 trials and saves the posterior columns required for assessment.
The batch is restartable: existing trial files are reused unless
`overwrite = TRUE`.

```{r fit-loo-models, eval=FALSE}
loo_paths <- fit_loo_historical_models(
  data = trial_sim_dat,
  output_dir = "data-raw/loo-random-inverse-gamma",
  seed = 2026,
  nchains = 4,
  ncores = 4,
  niter = 2000,
  nwarmup = 1000,
  adapt_delta = 0.95,
  max_treedepth = 15
)
```

`loo_cv_model_assessment()` calculates two quantities:

- LOO-CV RMSE compares each held-out clinical estimate with its posterior mean
  from the full-carryover MBE update.
- Tolerance coverage predicts the held-out observed clinical estimate after
  conditioning on only its two observed surrogate estimates.

For posterior draw \(b\), the predictive clinical estimate has distribution

$$
N\left(
  \beta_0^{(b)} + \beta_1^{(b)}\gamma_{01}^{(b)} +
  \beta_2^{(b)}\gamma_{02}^{(b)},
  \lambda_{\theta}^{2(b)} + \sigma^2_{\hat{\theta}_0}
\right).
$$

The variance therefore includes both the historical model's residual clinical
heterogeneity and the held-out trial's clinical sampling variance.

```{r calculate-loo-assessment, eval=FALSE}
loo_assessment <- loo_cv_model_assessment(
  loo_posteriors = "data-raw/loo-random-inverse-gamma",
  data = trial_sim_dat,
  seed = 2026
)
```

The completed, converged 66-trial assessment is included as two compact package
datasets so users do not need to rerun all 66 Stan models:

```{r packaged-loo-results}
data("loo_assessment_summary", package = "MBE")
data("loo_assessment_by_trial", package = "MBE")

knitr::kable(loo_assessment_summary, digits = 4)

coverage_counts <- data.frame(
  interval = c("95%", "90%"),
  covered = c(
    sum(loo_assessment_by_trial$covered_95),
    sum(loo_assessment_by_trial$covered_90)
  ),
  total = nrow(loo_assessment_by_trial)
)
knitr::kable(coverage_counts)
```

Only these small derived tables are installed with MBE. The complete
leave-one-out posterior files remain in `data-raw/` or suitable external
archival storage.
