---
title: "Mapping the MBE Manuscript to the Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping the MBE Manuscript to the Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MBE)
data("trial_sim_dat", package = "MBE")
```

## Overview

The Multi-Component Bayesian Endpoint (MBE) workflow described by Lee et al.
(2026) has two stages:

1. Fit a trial-level Bayesian meta-analytic model to historical randomized
   clinical trials.
2. Use draws from that historical model to update the treatment effects for a
   new trial.

The package implementation is deliberately restricted to one clinical endpoint
and exactly two surrogate endpoints. In the chronic kidney disease example, the
two surrogates are treatment effects on chronic and acute eGFR slopes.

## Input data

The historical dataset should be either a `data.table` or `data.frame` with the
columns names exactly as described below.:

| Column | Meaning |
|:--|:--|
| `CE_est` | Estimate of the treatment effect on the clinical endpoint |
| `CE_se` | Standard error of the estimated treatment effect on the clinical endpoint |
| `Sur1_est` | Estimate of the treatment effect on the first surrogate endpoint |
| `Sur1_se` | Standard error of the estimated treatment effect on the first surrogate endpoint |
| `Sur2_est` | Estimate of the treatment effect on the second surrogate endpoint |
| `Sur2_se` | Standard error of the estimated treatment effect on the second surrogate endpoint |
| `Cor_CE_Sur1` | Sampling correlation between `CE_est` and `Sur1_est` |
| `Cor_CE_Sur2` | Sampling correlation between `CE_est` and `Sur2_est` |
| `Cor_Sur1_Sur2` | Sampling correlation between `Sur1_est` and `Sur2_est` |

`trial_sim_dat` provides a shareable simulated example with 66 trials. The
restricted CKD-EPI CT trial data used in the motivating analysis are not
distributed with the package.

## Stage 1: fit the historical model

The historical model can use a fixed or estimated clinical intercept and either
half-normal or inverse-gamma priors for its uncertainty parameters. The
manuscript's initial specification uses an random intercept $\beta_0$ and
inverse-gamma priors on the variance parameters.  The first simulated trial can be held out as follows:

```{r historical-fit, eval=FALSE}
historical_fields <- c(
  "CE_est", "CE_se", "Sur1_est", "Sur1_se", "Sur2_est", "Sur2_se",
  "Cor_CE_Sur1", "Cor_CE_Sur2", "Cor_Sur1_Sur2"
)
historical_fit <- historical_model_fit_2surrogates(
  data = as.data.frame(trial_sim_dat)[-1, historical_fields],
  random_intercept = TRUE, #allow for a random intercept. Set to FALSE to fix the intercept at zero.
  prior_for_uncertainty = "inverse_gamma", #prior for the variance parameters. If set as "half_normal", then half normal distribution will be modeled on the standard deviation parameters.
  nchains = 4, #Number of chains
  ncores = 4, #Number of cores for parallel computation. One core per chain is recommended. If set to 1, then all chains will be run sequentially on a single core.
  niter = 2000, #Number of post-warmup samples to save for each chain
  nwarmup = 1000, #Number of warmup samples to discard for each chain
  seed = 2026,
  adapt_delta = 0.95,
  max_treedepth = 15
)

historical_fit$summary
historical_fit$loo
historical_fit$waic
```

The returned object also contains the CmdStan fit, posterior draws, and sampler
diagnostics. Historical fits should be accepted only after checking convergence
and sampler warnings.

## Stage 2: update the MBE for a new trial

The package includes `historical_posterior`, a compact set of compatible draws
from the random-intercept, inverse-gamma model with trial 1 held out.

```{r historical-posterior}
data("historical_posterior", package = "MBE")
head(historical_posterior)
```

The column names of this posterior datasets are different from what's used in the manuscript. Here's the mapping:

| Manuscript name | Programming name |
|:--|:--|
| $\beta_0$ | `alphaCEonSur1Sur2` |
| $\beta_1$ | `b1CEonSur1Sur2` |
| $\beta_2$ | `b2CEonSur1Sur2` |
| $\lambda^2_\theta$ | `SigSqCEonSur1Sur2` |
| $\alpha_0$ | `alphaSur1onSur2` |
| $\alpha_1$ | `bSur1onSur2` |
| $\lambda^2_\gamma$ | `SigSqSur1onSur2` |
| $\mu_{\gamma 2}$ | `muSur2` |
| $\sigma^2_{\gamma 2}$ | `sigSqSur2` |

This posterior distribution is used as prior for estimating the treatment effect on the clinical endpoint for the held-out trial. The held-out trial is converted to the naming convention used by `MBE()`:

```{r mbe-update}
new_trial <- list(
  ClnEst = trial_sim_dat$CE_est[1],
  ClnSE = trial_sim_dat$CE_se[1],
  Sur1Est = trial_sim_dat$Sur1_est[1],
  Sur1SE = trial_sim_dat$Sur1_se[1],
  Sur2Est = trial_sim_dat$Sur2_est[1],
  Sur2SE = trial_sim_dat$Sur2_se[1],
  R1Clin = trial_sim_dat$Cor_CE_Sur1[1],
  R2Clin = trial_sim_dat$Cor_CE_Sur2[1],
  R12 = trial_sim_dat$Cor_Sur1_Sur2[1]
)

set.seed(1)
mbe_fit <- MBE(
  mcmc_dat = historical_posterior,
  sample_dat = new_trial,
  diffuse = TRUE,
  diffuse_se = 100,
  intercept0 = TRUE
)

mbe_fit$post_mean
mbe_fit$post_var
mbe_fit$post_quantiles_clinical
mbe_fit$importance_diagnostics
```

`post_mean` and `post_var` are the mean and covariance of the updated posterior
mixture. These quantities are calculated using weighted sampling method shown in 
Appendix B of the manuscript. `post_psi0` contains resampled posterior draws in the fixed order
clinical endpoint, surrogate 1, and surrogate 2. `weight_data$norm_w` is the set of normalized weights, $\{ \tilde{w}^{(b)} \}$. `importance_diagnostics` helps identify
whether those weights are concentrated in only a small number of historical
draws.

## Historical-model fit assessment

Fitting and assessment are separate steps. Choose the historical-model
intercept and uncertainty prior, then fit the model to all but one trial using
`historical_model_fit_2surrogates()`. Repeat for each held-out trial. The two
assessment functions, `loo_cv_rmse()` and `tolerance_interval_coverage()`, only
summarize inputs supplied by the user; they never fit models or select priors.

This is internal leave-one-trial-out assessment: each held-out trial serves as
a pseudo-new trial. The package uses simulated trials, so its results are not
expected to match the manuscript's CKD trial results.

### Prepare the fits and point estimates yourself

The following example shows the fitting loop explicitly. The settings below
are illustrative and can be changed independently of the assessment functions.
`niter` is the number of saved post-warmup draws per chain. Run this expensive
step only when new fits are needed; it is not run when building the vignette.

```{r fit-loo-models, eval=FALSE}
historical_fields <- c(
  "CE_est", "CE_se", "Sur1_est", "Sur1_se", "Sur2_est", "Sur2_se",
  "Cor_CE_Sur1", "Cor_CE_Sur2", "Cor_Sur1_Sur2"
)
trials <- as.data.frame(trial_sim_dat)
historical_data <- trials[, historical_fields]

# Your choices for this analysis; the metric functions do not set these.
use_random_intercept <- TRUE
uncertainty_prior <- "inverse_gamma"
use_diffuse_surrogate_prior <- FALSE
center_clinical_intercept <- FALSE

paired_effects <- data.frame(
  observed = trials$CE_est,
  estimated = rep(NA_real_, nrow(trials)),
  row.names = as.character(trials$trial_id)
)
loo_draws <- vector("list", nrow(trials))
names(loo_draws) <- as.character(trials$trial_id)
loo_diagnostics <- vector("list", nrow(trials))

for (i in seq_len(nrow(trials))) {
  historical_fit <- historical_model_fit_2surrogates(
    data = historical_data[-i, , drop = FALSE],
    random_intercept = use_random_intercept,
    prior_for_uncertainty = uncertainty_prior,
    nchains = 4, ncores = 4, niter = 2000, nwarmup = 1000,
    seed = 2026 + i - 1,
    adapt_delta = 0.95, max_treedepth = 15
  )
  loo_draws[[i]] <- historical_fit$posterior_samples
  loo_diagnostics[[i]] <- list(
    sampler = historical_fit$diagnostics,
    convergence = historical_fit$fit$summary()
  )

  held_out <- list(
    ClnEst = trials$CE_est[i], ClnSE = trials$CE_se[i],
    Sur1Est = trials$Sur1_est[i], Sur1SE = trials$Sur1_se[i],
    Sur2Est = trials$Sur2_est[i], Sur2SE = trials$Sur2_se[i],
    R1Clin = trials$Cor_CE_Sur1[i],
    R2Clin = trials$Cor_CE_Sur2[i], R12 = trials$Cor_Sur1_Sur2[i]
  )
  set.seed(2026 + i - 1)
  updated <- MBE(
    mcmc_dat = loo_draws[[i]], sample_dat = held_out,
    diffuse = use_diffuse_surrogate_prior,
    intercept0 = center_clinical_intercept
  )
  paired_effects$estimated[i] <- updated$post_mean[1, 1]
}

# Inspect diagnostics and resolve fitting issues before interpreting the metric.
loo_diagnostics[[1]]
loo_cv_rmse(paired_effects)
```

If you already have fits, reuse their posterior draws in the `MBE()` calls;
there is no need to repeat the historical fitting step. Saved records from the
earlier development workflow can still be read with
`readRDS(path)$posterior_draws`.

For example, if the original 66 saved records are on your computer, load them
directly instead of running the fitting loop:

```{r reuse-saved-fits, eval=FALSE}
trials <- as.data.frame(trial_sim_dat)
loo_draws <- lapply(as.character(trials$trial_id), function(id) {
  record <- readRDS(file.path(
    "data-raw/loo-random-inverse-gamma", paste0("heldout_trial_", id, ".rds")
  ))
  stopifnot(identical(as.character(record$held_out_trial), id))
  record$posterior_draws
})
names(loo_draws) <- as.character(trials$trial_id)
use_diffuse_surrogate_prior <- FALSE
center_clinical_intercept <- FALSE
```

### Calculate RMSE from a two-column table

Supply one row per trial and two columns named `observed` and `estimated`.
A data frame or numeric matrix is accepted, with columns in either order.
Match trial IDs before assembling the pairs. The estimate is chosen by the
user; the example uses the posterior mean from `MBE()`, an estimate of the true
effect rather than a known true effect.

```{r rmse-from-saved-results}
data("loo_assessment_by_trial", package = "MBE")
paired_effects <- data.frame(
  observed = loo_assessment_by_trial$observed_clinical,
  estimated = loo_assessment_by_trial$posterior_mean_clinical
)
loo_cv_rmse(paired_effects)
loo_cv_rmse(as.matrix(paired_effects))
```

The calculation is `sqrt(mean((observed - estimated)^2))`. With a full MBE
update, the held-out clinical estimate contributes to its own posterior mean.
This measures agreement after updating, rather than prediction from surrogates
alone. Calling the metric LOO-CV assumes the corresponding trial was excluded
from historical fitting; the metric function cannot verify that step.

### Generate the clinical predictive distribution and check capture

`clinical_predictive_distribution()` takes historical posterior draws and the
held-out trial's surrogate estimates, standard errors, and sampling correlations.
It generates draws for the observed clinical estimate conditional on the two
surrogate estimates. The clinical estimate itself is ignored even if `ClnEst`
is present in the input list. This example reuses `historical_posterior` and
`new_trial` from above; trial 1 was excluded from that historical fit:

```{r clinical-prediction}
predictive_draws <- clinical_predictive_distribution(
  mcmc_dat = historical_posterior, sample_dat = new_trial,
  diffuse = FALSE, intercept0 = FALSE,
  n_draws = 4000, seed = 2026
)
stats::quantile(predictive_draws, c(0.025, 0.975))
```

`tolerance_interval_coverage()` calls this prediction function internally. Supply
the desired lower and upper quantile probabilities: `c(0.025, 0.975)` for a 95%
interval, or one row per interval in a two-column matrix. Asymmetric intervals
are also supported.

```{r interval-capture}
interval_probs <- rbind(
  c(0.025, 0.975),  # 95%
  c(0.05, 0.95),    # 90%
  c(0.075, 0.925)   # 85%
)
tolerance_interval_coverage(
  observed = new_trial$ClnEst,
  mcmc_dat = historical_posterior, sample_dat = new_trial,
  probs = interval_probs,
  diffuse = FALSE, intercept0 = FALSE,
  n_draws = 4000, seed = 2026
)
```

The result has one row per probability pair: `lower_prob`, `upper_prob`,
`level` (as a percentage), `observed`, `lower`, `upper`, and logical `covered`.
Bounds use `stats::quantile(type = 7)`, and both endpoints count as covered.
The function generates predictive draws once per trial, then uses the same
draws for all requested intervals. A fixed seed makes simulation reproducible.

For multiple trials, reuse the `loo_draws` list from the fitting example,
match it to the trial IDs, and combine single-trial results:

```{r aggregate-coverage, eval=FALSE}
stopifnot(
  !is.null(names(loo_draws)), !anyDuplicated(names(loo_draws)),
  !anyDuplicated(trials$trial_id),
  setequal(as.character(trials$trial_id), names(loo_draws))
)
loo_draws <- loo_draws[as.character(trials$trial_id)]
capture <- do.call(rbind, lapply(seq_len(nrow(trials)), function(i) {
  held_out <- list(
    ClnSE = trials$CE_se[i],
    Sur1Est = trials$Sur1_est[i], Sur1SE = trials$Sur1_se[i],
    Sur2Est = trials$Sur2_est[i], Sur2SE = trials$Sur2_se[i],
    R1Clin = trials$Cor_CE_Sur1[i],
    R2Clin = trials$Cor_CE_Sur2[i], R12 = trials$Cor_Sur1_Sur2[i]
  )
  result <- tolerance_interval_coverage(
    observed = trials$CE_est[i],
    mcmc_dat = loo_draws[[i]], sample_dat = held_out,
    probs = interval_probs, n_draws = 8000, seed = 2026 + i - 1,
    diffuse = use_diffuse_surrogate_prior,
    intercept0 = center_clinical_intercept
  )
  data.frame(trial_id = trials$trial_id[i], result)
}))
stats::aggregate(covered ~ lower_prob + upper_prob, data = capture, FUN = mean)
```

Prediction conditions the joint observed-effect distribution on the two
surrogate estimates. It incorporates residual clinical heterogeneity, clinical
sampling variance, surrogate uncertainty, and the supplied sampling correlations.
Historical draws are weighted by the likelihood of the observed surrogates.
This is different from passing latent-effect draws from `MBE()$post_psi0`,
which already condition on the observed clinical estimate.

### Archived example results

The completed 66-trial assessment remains available as two compact datasets.
Their provenance is the random-intercept, inverse-gamma specification; neither
new metric requires that specification. The archived coverage used an empirical
percentile rule and an unweighted surrogate-update simulation. The new function
uses joint conditioning, surrogate-based weights, and inclusive quantile bounds.
Those differences can change coverage. Archived results are preserved, not
recalculated or relabeled as results of the new prediction function.

```{r packaged-loo-results}
data("loo_assessment_summary", package = "MBE")
data("loo_assessment_by_trial", package = "MBE")

knitr::kable(loo_assessment_summary, digits = 4)

coverage_counts <- data.frame(
  interval = c("95%", "90%"),
  covered = c(
    sum(loo_assessment_by_trial$covered_95),
    sum(loo_assessment_by_trial$covered_90)
  ),
  total = nrow(loo_assessment_by_trial)
)
knitr::kable(coverage_counts)
```

Only these small derived tables are installed with MBE. The complete
leave-one-out posterior files remain in `data-raw/` or suitable external
archival storage.
