---
title: "Simulation Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MBE)
data("trial_sim_dat", package = "MBE")
data("historical_posterior", package = "MBE")
```

## Simulated trial data

`trial_sim_dat` contains 66 simulated trials with one clinical endpoint and
exactly two surrogate endpoints. Each row contains treatment-effect estimates,
standard errors, sampling correlations, and the corresponding simulation truths.

```{r trial-data}
head(trial_sim_dat)
```

The column definitions are available with `?trial_sim_dat`.

## Fit the historical model

For this example, trial 1 is treated as a new trial and the historical model is
fit to the remaining 65 trials. The call below uses the manuscript's initial
random-intercept, inverse-gamma specification. It is not run while building the
vignette because it requires CmdStan and several minutes of computation.

```{r historical-model, eval=FALSE}
historical_fit <- historical_model_fit_2surrogates(
  data = trial_sim_dat[-1, ],
  random_intercept = TRUE,
  prior_for_uncertainty = "inverse_gamma",
  nchains = 4,
  ncores = 4,
  niter = 2000,
  nwarmup = 1000,
  seed = 2026,
  adapt_delta = 0.95,
  max_treedepth = 15
)

historical_fit$diagnostics
historical_fit$summary

historical_posterior <- as.data.frame(historical_fit$posterior_samples)[, c(
  "alphaCEonSur1Sur2",
  "b1CEonSur1Sur2",
  "b2CEonSur1Sur2",
  "SigSqCEonSur1Sur2",
  "alphaSur1onSur2",
  "bSur1onSur2",
  "SigSqSur1onSur2",
  "muSur2",
  "sigSqSur2"
)]
```

The package's `historical_posterior` dataset contains compatible precomputed
draws from this specification with trial 1 held out.

## Update the MBE

Construct the new-trial input from the held-out row:

```{r new-trial}
new_trial <- list(
  ClnEst = trial_sim_dat$CE_est[1],
  ClnSE = trial_sim_dat$CE_se[1],
  Sur1Est = trial_sim_dat$Sur1_est[1],
  Sur1SE = trial_sim_dat$Sur1_se[1],
  Sur2Est = trial_sim_dat$Sur2_est[1],
  Sur2SE = trial_sim_dat$Sur2_se[1],
  R1Clin = trial_sim_dat$Cor_CE_Sur1[1],
  R2Clin = trial_sim_dat$Cor_CE_Sur2[1],
  R12 = trial_sim_dat$Cor_Sur1_Sur2[1]
)
```

Then update the historical distribution with the new-trial estimates:

```{r mbe-estimation}
set.seed(1)
mbe_fit <- MBE(
  mcmc_dat = historical_posterior,
  sample_dat = new_trial,
  diffuse = TRUE,
  diffuse_se = 100,
  intercept0 = TRUE
)

mbe_fit$post_mean
mbe_fit$post_var
mbe_fit$post_quantiles_clinical
mbe_fit$importance_diagnostics
```

The rows of `post_mean` and `post_var` are ordered as the clinical endpoint,
surrogate 1, and surrogate 2. The posterior is a mixture over historical-model
draws; `post_mean` and `post_var` summarize its first two moments rather than
asserting that the mixture itself is a single multivariate normal distribution.

Posterior samples and normalized importance weights are also available:

```{r posterior-draws}
head(mbe_fit$post_psi0)
head(mbe_fit$weight_data[, c("clinical", "surrogate1", "surrogate2", "norm_w")])
```
