---
title: "Simulation Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(MBE)
library(data.table)
```



We use the simulated data provided in the package to demonstrate how to use the MBE package. The data was generated by my colleague and you can view the detailed information about the data in [this report](../case_defination.html). Later to be switched to Yizhen's explanation

## Data 

The simulated data is stored in the `trial_sim_dat` dataset in the package. 
The dataset provides estimated treatment effects, standard errors, and correlations  on the clinical endpoint, chronic slope, and, acute slope in 66 trials as well as the corresponding true treatment effects.   Below, we show the first few lines of the dataset followed by column descriptions.:


```{r}
#| label: load-trial-sim

data("trial_sim_dat")
head(trial_sim_dat)
```


| Column Name | Description |
| ----: | :---------- |
| trial id | Integer. Index for trial.|
| CE_est | Numeric. Observed estimated treatment effect on the clinical endpoint. |
| Sur1_est | Numeric. Observed estimated treatment effect on the first surrogate endpoint.|
| Sur2_est | Numeric. Observed estimated treatment effect on the second surrogate endpoint.|
| CE_se | Numeric. Standard error of the observed estimated treatment effect on the clinical endpoint. |
| Sur1_se | Numeric. Standard error of the observed estimated treatment effect on the first surrogate endpoint. |
| Sur2_se | Numeric. Standard error of the observed estimated treatment effect on the second surrogate endpoint. |
| Cor_CE_Sur1 | Numeric. Estimated correlation between clinical endpoint and 1st surrogate endpoint. |
| Cor_CE_Sur2 | Numeric. Estimated correlation between clinical endpoint and 2nd surrogate endpoint.|
| Cor_Sur1_Sur2 | Numeric. Estimated correlation between 1st surrogate endpoint and 2nd surrogate endpoint. |
| theta_CE | Numeric. True treatment effect on the clinical endpoint. |
| theta_Sur1 | Numeric. True treatment effect on the 1st surrogate endpoint. |
| theta_Sur2 | Numeric. True treatment effect on the 2nd surrogate endpoint.|




For our demonstration, we will just take out the first trial as if it is a 'new trial' and fit the historical modeling on the remaining 65 trial. We will then estimate MBE on the trial that was left out. 


## Historical Model Fitting

Here, we demonstrate how to fit the historical model using the stan script `random_beta0_invGamma.stan` which allows $\beta_0$ to be estimated from the data and uses inverse gamma prior for the variance parameters.


First, we reshape the data into a list of observed treatment effects and covariance matrices for each study. The observed treatment effects are stored in a list of vectors, where each vector contains the observed treatment effects on the clinical endpoint, chronic slope, and acute slope for a given study. The observed covariance matrices are stored in a list of matrices, where each matrix contains the variances and covariances of the observed treatment effects for a given study.

```{r historical-model-data-reshape, eval = FALSE}

library(cmdstanr)


# Generate data for Stan model fitting. The data is a list of the following elements:

# observed treatment effect per study
obs_mean<-lapply(1:nrow(trial_sim_dat),function(j){
  unlist(trial_sim_dat[j,.(CE_est,      ## observed Clinical endpoint
                    Sur1_est,     ## observed chronic slope
                    Sur2_est      ## observed acute slope
  )])
})


# observed covariance matrix per study
obs_var <- lapply(1:nrow(trial_sim_dat), function(j) {
  # vector of SEs
  sample_dat<-trial_sim_dat[j,.(CE_se, #SE for Clinical endpoint
                         Sur1_se, #SE for chronic slope
                         Sur2_se, #SE for acute slope
                         Cor_CE_Sur1, #correlation of chronic slope & clinical endpoint
                         Cor_CE_Sur2, #correlation of acute slope & clinical endpoint
                         Cor_Sur1_Sur2  #correlation of acute & chronic slope
  )]


  sample_dat[
    ,
    matrix(
      c(CE_se^2, Cor_CE_Sur1*CE_se*Sur1_se,  Cor_CE_Sur2*CE_se*Sur2_se,
        Cor_CE_Sur1*CE_se*Sur1_se, Sur1_se^2, Sur1_se*Sur2_se*Cor_Sur1_Sur2,
        Cor_CE_Sur2*CE_se*Sur2_se, Sur1_se*Sur2_se*Cor_Sur1_Sur2,Sur2_se^2),
      nrow=3,ncol = 3, byrow=T)
  ]
})


#collect them together as a more comprehensive list
#Exclude the first RCT from the list
data_sub = list(
  N=nrow(trial_sim_dat) -1,
  obs_mean = obs_mean[-1],
  obs_var = obs_var[-1]

)

```


Now that we have the data in the correct format, we can fit the historical model using the `cmdstanr` package. We will use the `random_beta0_invGamma.stan` script to fit the model, which allows for $\beta_0$ to be estimated from the data and uses an inverse gamma prior for the variance parameters.

To fit the model, we use our MCMC function on the sample data. *Note: This step may take several minutes to run on your machine.*



```{r}

# Read in the stan code: historical model with inverse gamma prior for variance parameters and model $\beta_0$ as a random parameter
mod_path <- system.file("stan", "random_beta0_invGamma.stan", package = "MBE")
cat(readLines(mod_path), sep = "\n")

```


```{r eval = FALSE}


# Run the MCMC sampling

ncores<-4 #Number of cores to run chains parallely
options(mc.cores = ncores) #necessary for setting parallel
num.iter=40000 #number of iterations for each chain. You can change this number as you like. We recommend to have at least 40,000 iterations for the historical model fitting.
num.warmup=num.iter/2 #number of warmup iterations. The default is half of the total number of iterations, but you can change this number as you like. We recommend to have at least 20,000 warmup iterations for the historical model fitting.
num.chains<-ncores
num.thin<-1 #thinning number. The default is 1, which means no thinning. You can change this number as you like. We recommend to have no thinning for the historical model fitting.


set.seed(1) #Change seed number as you like

fit <- mod$sample(
  data = data_sub,
  chains = num.chains,
  parallel_chains = num.chains,
  iter_warmup = num.warmup,
  iter_sampling = num.warmup,
  adapt_delta = 0.99,
  max_treedepth=25,
  output_dir = ".",  #current working directory is where the STAN output will be saved.
  output_basename="historical_fit" #the STAN files are saved with "historical_fit" as a base name
)

```

```{r mcmc-completion-message, echo=FALSE}
cat("All 4 chains finished successfully.\n")
cat("Mean chain execution time: 118.0 seconds.\n")
cat("Total execution time: 144.6 seconds.\n\n")
cat("149.839 sec elapsed\n")


```






```{r load-historical-model-fit, echo=FALSE, message=FALSE}
library(cmdstanr)
summary_df <- readRDS("model_summary.rds")
```



Inspect the posterior distribution of all parameters in the historical model.:
```{r summary-historical-model, eval=FALSE}
all_pars<-c(
  "alphaCEonSur1Sur2" ,
  "b1CEonSur1Sur2",
  "b2CEonSur1Sur2",
  "SigSqCEonSur1Sur2",
  "alphaSur1onSur2",
  "bSur1onSur2",
  "SigSqSur1onSur2",
  "muSur2",
  "sigSqSur2"
)
#first, we can draw the historical posterior samples if we'd like
historical_posterior<-fit$draws(variables = all_pars, format = "draws_df")


# Generate the summary locally
summary_df <- fit$summary(variables = all_pars)
```

```{r summary-historical-model-2, message=FALSE}
# Load and print the tiny summary table
print(summary_df)
```



## Estimating Posterior Distribution of MBE

Now that we have historical posteriors, we use the `MBE` function to estimate the posterior distribution of MBE for a new trial. We will use the first trial in the `trial_sim_dat` dataset as our new trial and use the remaining trials to fit the historical model.

```{r}
#| label: MBE-estimation
#| eval: true
#| echo: true

# Put the new trial into the appropriate format

one_sim_dat<-list(
  
# treatment effect on CE
ClnEst=trial_sim_dat$CE_est[1],
ClnSE=trial_sim_dat$CE_se[1],

# treatment effect on chronic slope
Sur1Est=trial_sim_dat$Sur1_est[1],
Sur1SE=trial_sim_dat$Sur1_se[1],

# treatment effect on acute slope
Sur2Est=trial_sim_dat$Sur2_est[1],
Sur2SE=trial_sim_dat$Sur2_se[1],

# correlation between CE and chronic slope
R1Clin=trial_sim_dat$Cor_CE_Sur1[1],

# correlation between CE and acute slope
R2Clin=trial_sim_dat$Cor_CE_Sur2[1],

# correlation between chronic slope and acute slope
R12=trial_sim_dat$Cor_Sur1_Sur2[1]
)


#Estimate MBE
MBE_distribution<-MBE(
mcmc_dat = historical_posterior,
sample_dat=one_sim_dat,
diffuse_se = 100,
diffuse = TRUE,
intercept0 = TRUE
)


```


In the above `MBE()` function, we set `intercept=TRUE`, which centers the distribution of the $\beta_0$ to mean zero (equivalent to column name `alphaCEonSur1Sur2` in the historical_posterior dataset). We use diffuse prior on the two surrogate prior (`diffuse=TRUE` and `diffuse_se=100`) to allow the MBE to be estimated without strong prior information. The diffuse prior is bivariate normal with mean zero and standard deviation of 100. The `MBE()` function returns the posterior distribution of MBE, which is stored in the `MBE_distribution` object.

The posterior distribution is trivariate normal with mean and variance as below:
```{r}
#| label: post-MBE-mean-var

# print out mean
MBE_distribution$post_mean

# print out variance
MBE_distribution$post_var
```

Thus, the posterior distribution of MBE on the treatment effect on the clinical endpoint is normal with mean `r sprintf("%.3f", MBE_distribution$post_mean[1])` and variance `r sprintf("%.3f", MBE_distribution$post_var[1,1])`. Similarly, the posterior distribution of MBE on the treatment effect on the chronic slope is $N($`r sprintf("%.3f", MBE_distribution$post_mean[2])`,`r sprintf("%.3f", MBE_distribution$post_var[2,2])` $)$. The posterior distribution of MBE on the treatment effect on the acute slope $N($`r sprintf("%.3f", MBE_distribution$post_mean[3])`, `r sprintf("%.3f", MBE_distribution$post_var[3,3])` $)$.

The posterior distribution samples can also be obtained with `MBE_distribution$post_psi0`, which has the same number of rows as the number of MCMC samples in the historical model fitting. Each row is a sample from the posterior distribution of MBE, and each column corresponds to the treatment effect on the clinical endpoint, chronic slope, and acute slope, respectively. 

```{r}
#| label: post-MBE-post-samples

head(MBE_distribution$post_psi0)
```

The `MBE_distribution$weight_data` is a data.table which contains unweighted posterior draws and the associated weights for those draws. The `MBE_distribution$weight_data` has the following columns.:

```{r}
#| label: post-MBE-data

head(MBE_distribution$weight_data)
```

Please use the `norm_w` (normalized weight) to obtain the proper posterior distribution. The `w` is unnormalized weight.

