---
title: "Simulation Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(MBE)
library(data.table)
```



We use the simulated data provided in the package to demonstrate how to use the MBE package. The data was generated by my colleague and you can view the detailed information about the data in [this report](../case_defination.html).

## Data 

The simulated data is stored in the `sim_dat` dataset in the package. 
The dataset provides treatment effects on the clinical endpoint (CE), chronic slope (Sur1), 
acute slope (Sur2), and log UACR (Sur3) estimated monthly from month 7 until 72,  
under 6 different scenarios.  Below, we show the first few lines of othe dataset followed by column descriptions.:


```{r}
head(sim_dat)
```


| Column Name | Description |
| ----: | :---------- |
| case | The scenario case number (2, 4, or 6).|
| analysis.month | Number of months since the study began. |
| ClnEst | observed treatment effect on Clinical endpoint (CE) |
| ClnSE | standard error of observed treatment effect on CE |
| Sur1Est | observed treatment effect on chronic slope (Surrogate 1) |
| Sur1SE | standard error of observed treatment effect on chronic slope |
| Sur2Est | observed treatment effect on acute slope (Surrogate 2) |
| Sur2SE | standard error of observed treatment effect on acute slope |
| Sur3Est | observed treatment effect on log UACR (Surrogate 3) |
| Sur3SE | standard error of observed treatment effect on log UACR |
| R1Clin | correlation between observed treatment effect on CE and chronic slope |
| R2Clin | correlation between observed treatment effect on CE and acute slope |
| R3Clin | correlation between observed treatment effect on CE and log UACR |
| R12 | correlation between observed treatment effect on chronic slope and acute slope |
| R13 | correlation between observed treatment effect on chronic slope and log UACR |
| R23 | correlation between observed treatment effect on acute slope and log UACR |


The `NA`s shown in the dataset is due to missing estimates for the CE
for early month, where there was insufficient number of patients who experienced 
the CE and thus the treatment effect on the CE could not be estimated. 
We handle this by imputing the missing estimates for the CE using 0 for the `ClnEst` 
and a large number (100) for the `ClnSE`. The correlation between the CE and the surrogates are 
imputed with the mean of the observed correlations (assuming the correlation is constant over time).

```{r}

#impute missing CE estimates
sim_dat[is.na(ClnEst), ClnEst := 0] 
sim_dat[is.na(ClnSE), ClnSE := 100]


#impute missing correlations with mean of observed correlations
sim_dat[is.na(R1Clin), R1Clin := mean(R1Clin, na.rm = TRUE)]
sim_dat[is.na(R2Clin), R2Clin := mean(R2Clin, na.rm = TRUE)]
sim_dat[is.na(R3Clin), R3Clin := mean(R3Clin, na.rm = TRUE)]

```



For our demonstration, we will use months 7-24 as the historical data from cases 1-3, 5, and 6.
24 months is a usual length for GFR based phase 2/3 CKD trials. 

We will use case 4 as the new trial and estimate the posterior distribution of MBE at each month 
and evaludate the perofrmance. 


## Historical Model Fitting

Here, we demonstrate how to fit the historical model using the stan script `random_beta0_invGamma.stan` which allows $\beta_0$ to be estimated from the data and uses inverse gamma prior for the variance parameters.


First, we reshape the data into a list of observed treatment effects and covariance matrices for each study. The observed treatment effects are stored in a list of vectors, where each vector contains the observed treatment effects on the clinical endpoint, chronic slope, and acute slope for a given study. The observed covariance matrices are stored in a list of matrices, where each matrix contains the variances and covariances of the observed treatment effects for a given study.

```{r historical-model-data-reshape, eval = FALSE}

library(cmdstanr)


# We use cases 1-3, 5, and 6 as the historical data.
# Just fit for up to month 24.
myData<-myData[case %in% c(1,2,3,5,6) & analysis.month <= 24,]


#impute missing CE estimates
myData[is.na(ClnEst), ClnEst := 0]
myData[is.na(ClnSE), ClnSE := 100]


#impute missing correlations with mean of observed correlations
myData[is.na(R1Clin), R1Clin := mean(myData$R1Clin, na.rm = TRUE)]
myData[is.na(R2Clin), R2Clin := mean(myData$R2Clin, na.rm = TRUE)]
myData[is.na(R3Clin), R3Clin := mean(myData$R3Clin, na.rm = TRUE)]



# Generate data for Stan model fitting. The data is a list of the following elements:

# observed treatment effect per study
obs_mean<-lapply(1:nrow(myData),function(j){
  unlist(myData[j,.(ClnEst,      ## observed Clinical endpoint
                    Sur1Est,     ## observed chronic slope
                    Sur2Est      ## observed acute slope
  )])
})


# observed covariance matrix per study
obs_var <- lapply(1:nrow(myData), function(j) {
  # vector of SEs
  sample_dat<-myData[j,.(ClnSE, #SE for Clinical endpoint
                         Sur1SE, #SE for chronic slope
                         Sur2SE, #SE for acute slope
                         R1Clin, #correlation of chronic slope & clinical endpoint
                         R2Clin, #correlation of acute slope & clinical endpoint
                         R12  #correlation of acute & chronic slope
  )]


  sample_dat[
    ,
    matrix(
      c(ClnSE^2, R1Clin*ClnSE*Sur1SE,  R2Clin*ClnSE*Sur2SE,
        R1Clin*ClnSE*Sur1SE, Sur1SE^2, Sur1SE*Sur2SE*R12,
        R2Clin*ClnSE*Sur2SE, Sur1SE*Sur2SE*R12,Sur2SE^2),
      nrow=3,ncol = 3, byrow=T)
  ]
})


#collect them together as a more comprehensive list
data_sub = list(
  N=nrow(myData), # set number of studies. should be 66.
  obs_mean = obs_mean,
  obs_var = obs_var

)

```


Now that we have the data in the correct format, we can fit the historical model using the `cmdstanr` package. We will use the `random_beta0_invGamma.stan` script to fit the model, which allows for $\beta_0$ to be estimated from the data and uses an inverse gamma prior for the variance parameters.

To fit the model, we use our MCMC function on the sample data. *Note: This step may take several minutes to run on your machine.*



```{r}

# Read in the stan code: historical model with inverse gamma prior for variance parameters and model $\beta_0$ as a random parameter
mod_path <- system.file("stan", "random_beta0_invGamma.stan", package = "MBE")
cat(readLines(mod_path), sep = "\n")

```


```{r eval = FALSE}


# Run the MCMC sampling

ncores<-4 #Number of cores to run chains parallely
options(mc.cores = ncores) #necessary for setting parallel
num.iter=40000 #number of iterations for each chain. You can change this number as you like. We recommend to have at least 40,000 iterations for the historical model fitting.
num.warmup=num.iter/2 #number of warmup iterations. The default is half of the total number of iterations, but you can change this number as you like. We recommend to have at least 20,000 warmup iterations for the historical model fitting.
num.chains<-ncores
num.thin<-1 #thinning number. The default is 1, which means no thinning. You can change this number as you like. We recommend to have no thinning for the historical model fitting.


set.seed(1) #Change seed number as you like

fit <- mod$sample(
  data = data_sub,
  chains = num.chains,
  parallel_chains = num.chains,
  iter_warmup = num.warmup,
  iter_sampling = num.warmup,
  adapt_delta = 0.99,
  max_treedepth=25,
  output_dir = ".",  #current working directory is where the STAN output will be saved.
  output_basename="historical_fit" #the STAN files are saved with "historical_fit" as a base name
)

```

```{r mcmc-completion-message, echo=FALSE}
cat("All 4 chains finished successfully.\n")
cat("Mean chain execution time: 973.1 seconds.\n")
cat("Total execution time: 1031.6 seconds.\n\n")
cat("Warning: 73 of 80000 (0.0%) transitions ended with a divergence.\n")
cat("See [https://mc-stan.org/misc/warnings](https://mc-stan.org/misc/warnings) for details.\n\n")
cat("1043.836 sec elapsed\n")
```






```{r load-historical-model-fit, echo=FALSE, message=FALSE}
library(cmdstanr)
summary_df <- readRDS("model_summary.rds")
```



Inspect the posterior distribution of all parameters in the historical model.:
```{r summary-historical-model, eval=FALSE}
all_pars<-c(
  "alphaCEonSur1Sur2" ,
  "b1CEonSur1Sur2",
  "b2CEonSur1Sur2",
  "SigSqCEonSur1Sur2",
  "alphaSur1onSur2",
  "bSur1onSur2",
  "SigSqSur1onSur2",
  "muSur2",
  "sigSqSur2"
)

# 1. Generate the summary locally
summary_df <- fit$summary(variables = all_pars)
```

```{r summary-historical-model-2, message=FALSE}
# Load and print the tiny summary table
print(summary_df)
```

## Estimating Posterior Distribution of MBE

