---
title: "Interim Analysis Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interim Analysis Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MBE)
data("interim_sim_dat", package = "MBE")
data("historical_posterior", package = "MBE")
```

`interim_sim_dat` contains simulated interim analyses for eight scenarios. Each
row has one clinical endpoint and exactly two surrogate endpoints: chronic
(`Sur1`) and acute (`Sur2`) eGFR slopes. Some early analyses have missing
clinical estimates because too few clinical events had occurred; those rows are
not yet eligible for an MBE update.

```{r interim-data}
new_trial_fields <- c(
  "ClnEst", "ClnSE",
  "Sur1Est", "Sur1SE",
  "Sur2Est", "Sur2SE",
  "R1Clin", "R2Clin", "R12"
)

interim_data <- as.data.frame(interim_sim_dat)
eligible <- stats::complete.cases(interim_data[, new_trial_fields])
table(eligible)
```

The first complete interim analysis can be updated using the historical
posterior supplied with the package:

```{r interim-mbe}
first_complete <- which(eligible)[1]
new_trial <- as.list(interim_data[first_complete, new_trial_fields])

set.seed(1)
interim_mbe <- MBE(
  mcmc_dat = historical_posterior,
  sample_dat = new_trial,
  diffuse = TRUE,
  diffuse_se = 100,
  intercept0 = TRUE
)

interim_data[first_complete, c("case", "analysis.month")]
interim_mbe$post_mean
interim_mbe$post_quantiles_clinical
```

For a sequence of interim analyses, repeat the construction of `new_trial` for
each complete row. The historical posterior can be reused, but the treatment
effect estimates, standard errors, and correlations must all correspond to the
same interim time point.
